PoVaTiM: A Possibilistic Valid-Time Model \\ \vspace*{0.04truein}

### 2012-06-05 ijcis-pos-model/sources/timeRepresentation.tex
last modifications

\begin{definition}
\label{def:ill-known-time-point}
\emph{Ill-known time point}\\
Consider a time domain $\T$, uncertainty about the values of the ill-known time point $X$ is given by the possibility distribution $\pi_X$:
\begin{equation}
\label{eq:ill-known-time-point}
\Pi(X) = \pi_X(t) \in \left[0,1\right] , t \in \T
\end{equation}
\end{definition}

@@

\begin{definition}
\label{def:ill-known-domain}
\emph{Domain for an ill-known time point}\\
Consider $\Pow(\T)$ the set of all the possibility distributions over $\T$, and the three fuzzy constants \emph{UNKNOWN} $=\left\lbrace 1/t, \forall t \in \T \right\rbrace$, \emph{UNDEFINED} $=\left\lbrace 0/t,\forall t \in \T \right\rbrace$ and \emph{NULL} $=\lbrace 1/$ \emph{UNKNOWN}, $1/$ \emph{UNDEFINED} $\rbrace$. The domain for an ill-known time point $X$ is given by: 
\begin{align}
\label{eq:ill-known-domain}
\Dom (X) =  \lbrace \Pow(\T) & \cup \mbox{\emph{UNKNOWN}} \\
\nonumber
&\cup \mbox{\emph{UNDEFINED}} \\
\nonumber
&\cup \mbox{\emph{NULL}} \rbrace
%
\end{align}
\end{definition}

\subsubsection{\label{subsubsec:ill-known-time-datatypes}Datatypes}
The datatypes for an ill-known time point $X$ are shown in Table \ref{tbl:time-point-types}.
\vglue13pt
%\begin{table}[htbp]
\tcap{Data types for a time point}
\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c p{2cm} p{2cm}}\\
\hline
1 & A single time point &  $1/x, x \in \T$\\
2 & A possibility distribution in the numeric domain & A fuzzy number or a fuzzy interval.\\
3 & An unknown value & \textbf{UNKNOWN}$= \lbrace1/t,\forall t \in \T \rbrace$\\
4 & An undefined value & \textbf{UNDEFINED}$= \lbrace 0/t,\forall t \in \T \rbrace$\\
5 & A null value & \textbf{NULL} $=\lbrace 1/$Unknown, 1/Undefined $\rbrace$\\
\hline\\
\end{tabular}
\label{tbl:time-point-types}
}

@@

\begin{definition}
\emph{Ill-known time interval}\\
An ill-known time interval denoted by $\left[X, Y\right]$ is a precise time interval whose boundaries are not precisely known. The ill-known interval is defined by two ill-known points, namely $X$,$Y$.
In order to know if all the points in the crisp time interval $I=\left[a, b\right]$ are within the boundaries of the ill-known time interval $\left[X, Y\right]$ we define the following two ill-known constraints $C_{1}, C_{2}$ 
\begin{eqnarray}
C_1\stackrel{\triangle}{=}\left(\geq,X\right)  \\
C_2\stackrel{\triangle}{=}\left(\leq,Y\right)
\end{eqnarray}
Note that the boolean operator applied to the constraints is $\bool=\wedge$. Hence, the construct of both constraints is $C_1 \wedge C_2$
This means that we want to know if all the points in the interval $I$ are greater or equal to $X$ and, at the same time, smaller or equal to $Y$.
\end{definition}

@@

\subsubsection{\label{subsubsec:open-interval}Open ill-known time intervals}
Quite often, the user may want to specify time intervals with open boundaries in one or both endpoints. Consider an ill-known time interval $\left[X, Y\right]$. Then it is possible to distinguish between the following two types of open intervals:
\begin{enumerate}
\item
\begin{definition}
\emph{Completely unknown time interval}: Both starting and ending points are unknown, therefore the whole interval is specified by the \emph{UNKNOWN} constant. Due to the representation, both endpoints, say $X$ and $Y$, are stored. Thus, the interval is stored as $[$ \emph{UNKNOWN}, \emph{UNKNOWN} $]$. 
\end{definition}

\item
\begin{definition}
\emph{Semi-open time interval}: Only one of the two ill-known endpoints for the time interval $\left[X, Y\right]$ is unknown. Note that due to the ill-known constraints $C_{1}, C_{2}$ it is not possible a representation like $[$ \emph{UNKNOWN} $, Y]$ or $[X, $ \emph{UNKNOWN} $]$. The solution adopted for this problem is explained in the following paragraph. 
\end{definition}
\end{enumerate}

\paragraph{Representation of semi-open time intervals}
As mentioned before, the problem resides in the representation of these kind of interval. In order to get a short representation for these intervals, two constants are defined: $FB$ \emph{From the Beginning}: it is only valid for the left value of the interval and $UC$ \emph{Until Changed} (this name is usual in the temporal database community, see ~\cite{Jensen1994}) which is only valid for the right value of the interval. Note that these two constants are aliases for a function that obtains both possibility and necessity measures.

@@

\vglue13pt
%\begin{table}[htbp]
\tcap{Relations for the $\Open(C)$ function.}
\centerline{\small Relations}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c c}\\
\hline
$R$ & $\Rp$ & $\Rn$\\ \hline
$<$ & $>$ & $\leq$ \\
$\leq$ & $\geq$ & $<$ \\
$>$ & $<$ & $\geq$ \\
$\geq$ & $\leq$ & $>$ \\
\hline\\
\end{tabular}
\label{tbl:open-pos-nec-rels}
}

@@

\begin{example}
Consider an ill-known time interval given by $\left[\FB, Y \right]$. Consider also that in this case, $Y= \left[15/10/2012,3,4 \right] $. Figure \ref{fig:example-open-time-interval} shows the representation for $Y$. The user wants to obtain the possibility and the necessity measures for the $\FB$ part of the interval are:
\end{example}

@@

\vglue13pt
%\begin{table}[htbp]
\tcap{Data types for an ill-known time interval $\left[X, Y\right]$.}
\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c p{2cm}}\\
\hline
$X$  & $Y$  & Description\\
\hline
$\lbrace1,2 \rbrace$ & $\lbrace1,2 \rbrace$ &  an ill-known time interval.\\
3 & 3 & An unknown time interval. \\
$\FB$ & $\lbrace1,2 \rbrace$ & a left-open time interval.\\
$\lbrace1,2 \rbrace$ & $\UC$ & a right-open time interval.\\
\hline\\
\end{tabular}
\label{tbl:time-interval-types}
}

\subsection{\label{subsec:fuzzy-allen-relations}Allen's relations}

\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:fuzzy-allen-relations}Constructs of constraints related to their respective Allen relations, as used in the presented work. In this table, the ill-known time interval $J = \left[X, Y\right]$ in a record $r$ has a start point described by possibilistic variable $X$ and an end point described by possibilistic variable $Y$. The crisp time interval in the user's temporal demand is denoted $I$.}
\centerline{\small Relations}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c p{4cm} }\\
\hline
Allen Relation & Construct of constraints\\
\hline
I before J & $C_1 \triangleq \left(<,X\right)$ \\
\hline
I equal J & $\left(C_1 \triangleq \left(\geq,X\right)\right) \wedge \neg \left(C_2 \triangleq \left(\neq,X\right) \right) \wedge \left(C_3 \triangleq \left(\leq,Y\right) \right) \wedge \neg \left(C_4 \triangleq \left(\neq,Y\right)\right)$ \\
\hline
I meets J & $\left(C_1 \triangleq \left(\leq,X\right)\right) \wedge \neg \left(C_2 \triangleq \left(\neq,X\right) \right)$ \\
\hline
I overlaps J & $\left(C_1 \triangleq \left(<,Y\right)\right) \wedge \neg \left(C_2 \triangleq \left(\leq,X\right) \right) \wedge \neg \left(C_3 \triangleq \left(\geq,X\right) \right)$ \\
\hline
I during J & $\left( \left (C_1 \triangleq \left(>,X\right) \right) \wedge \left(C_2 \triangleq \left(\leq,Y\right) \right) \right) \vee \left( \left(C_3 \triangleq \left(\geq,X\right) \right) \wedge  \left(C_4 \triangleq \left(<,Y\right) \right) \right)$ \\
\hline
I starts J & $\left(C_1 \triangleq \left(\geq,X\right) \right) \wedge \neg \left(C_2\triangleq \left(\neq,X\right)\right)$ \\
\hline
I finishes J & $\left(C_1 \triangleq \left(\leq,Y\right) \right) \wedge \neg \left(C_2\triangleq \left(\neq,Y\right)\right)$ \\
\hline\\
\end{tabular}
}

>>>>

\begin{definition}
\label{def:ill-known-time-point}
\textbf{Ill-known time point.}\\
Consider a time domain $\T$, uncertainty about the values of the ill-known time point $X$ is given by the possibility distribution $\pi_X$:
\begin{equation}
\label{eq:ill-known-time-point}
\Pi(X) = \pi_X(t) \in \left[0,1\right] , t \in \T
\end{equation}
\end{definition}

@@

\begin{definition}
\label{def:ill-known-domain}
\textbf{Domain for an ill-known time point.}\\
Consider $\Pow(\T)$ the set of all the possibility distributions over $\T$, and the three fuzzy constants \emph{UNKNOWN} $=\left\lbrace 1/t, \forall t \in \T \right\rbrace$, \emph{UNDEFINED} $=\left\lbrace 0/t,\forall t \in \T \right\rbrace$ and \emph{NULL} $=\lbrace 1/$ \emph{UNKNOWN}, $1/$ \emph{UNDEFINED} $\rbrace$. The domain for an ill-known time point $X$ is given by: 
\begin{align}
\label{eq:ill-known-domain}
\Dom (X) =  \lbrace \Pow(\T) & \cup \mbox{\emph{UNKNOWN}} \\
\nonumber
&\cup \mbox{\emph{UNDEFINED}} \\
\nonumber
&\cup \mbox{\emph{NULL}} \rbrace
%
\end{align}
\end{definition}

\subsubsection{\label{subsubsec:ill-known-time-datatypes}Datatypes}
The data type for the representation of an ill-known time point allows the representation of the values shown in Table \ref{tbl:time-point-types}
%The datatypes for an  $X$ are shown in Table \ref{}.
\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:time-point-types}Values for the time point data type.}
\centerline{\small TIME POINT DATA TYPE}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c p{0.4\columnwidth} p{0.3\columnwidth}}\\
Subtype & Value & Representation \\
\hline
1 & A single time point &  $1/x, x \in \T$\\
2 & A possibility distribution in the numeric domain & A fuzzy number or a fuzzy interval.\\
3 & An unknown value & \textbf{UNKNOWN}$= \lbrace1/t,\forall t \in \T \rbrace$\\
4 & An undefined value & \textbf{UNDEFINED}$= \lbrace 0/t,\forall t \in \T \rbrace$\\
5 & A null value & \textbf{NULL} $=\lbrace 1/$Unknown, 1/Undefined $\rbrace$\\
\hline\\
\end{tabular}
}

@@

\begin{definition}
\textbf{Ill-known time interval}\\
An ill-known time interval denoted by $\left[X, Y\right]$ is a precise time interval whose boundaries are not precisely known. The ill-known interval is defined by two ill-known points, namely $X$,$Y$.
In order to know if all the points in the crisp time interval $I=\left[a, b\right]$ are within the boundaries of the ill-known time interval $\left[X, Y\right]$ we define the following two ill-known constraints $C_{1}, C_{2}$ 
\begin{eqnarray}
\label{eq:constraint-c1}
C_1\stackrel{\triangle}{=}\left(\geq,X\right)  \\
\label{eq:constraint-c2}
C_2\stackrel{\triangle}{=}\left(\leq,Y\right)
\end{eqnarray}
Note that the boolean operator applied to the constraints is $\bool=\wedge$. Hence, the construct of both constraints is $C_1 \wedge C_2$
This means that we want to know if all the points in the interval $I$ are greater or equal to $X$ and, at the same time, smaller or equal to $Y$.
\end{definition}

@@

\subsubsection{\label{subsubsec:open-interval}Open ill-known time intervals}
Quite often, the user may want to specify time intervals with open boundaries in one or both endpoints. Consider an ill-known time interval $\left[X, Y\right]$. Then it is possible to distinguish between the following two types of open intervals:
%\begin{enumerate}
%\item
\begin{definition}
\textbf{Completely unknown time interval}: Both starting and ending points are unknown, therefore the whole interval is unknown.
%specified by the \emph{UNKNOWN} constant. Due to the representation, both endpoints, say $X$ and $Y$, are stored. Thus, the interval is stored as $[$ \emph{UNKNOWN}, \emph{UNKNOWN} $]$. 
\end{definition}

%\item
\begin{definition}
\textbf{Semi-open time interval}: Only one of the two ill-known endpoints for the time interval $\left[X, Y\right]$ is unknown. Example \ref{ex:open-time-interval} and Figure \ref{fig:example-open-time-interval} illustrates a left open time interval.
% Note that due to the ill-known constraints $C_{1}, C_{2}$ it is not possible a representation like $[$ \emph{UNKNOWN} $, Y]$ or $[X, $ \emph{UNKNOWN} $]$. The solution adopted for this problem is explained in the following paragraph. 
\end{definition}
%\end{enumerate}

\subsubsection{\label{subsubsec:representation-semi-open}Representation of semi-open time intervals}
As mentioned before, the problem resides in the representation of these kind of interval. In order to get a short representation for these intervals, two constants are defined: $FB$ \emph{From the Beginning}: it is only valid for the left value of the interval and $UC$ \emph{Until Changed} (this name is usual in the temporal database community, see ~\cite{Jensen1994}) which is only valid for the right value of the interval. Note that these two constants are aliases for a function that obtains both possibility and necessity measures.

@@

\vglue13pt
%\begin{table}[htbp]
\tcap{Relations for the $\Open(C)$ function. Depending on the relation $R = \left(\leq, <, >, \geq\right)$ in the constraint $C$, the values for $\Rp$ and $\Rn$ are shown.}
\centerline{\small Relations}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c c}\\
\hline
Constraint & $\Rp$ & $\Rn$\\ \hline
$C\stackrel{\triangle}{=}\left(<,T\right)$ & $>$ & $\leq$ \\
$C\stackrel{\triangle}{=}\left(\leq,T\right)$ & $\geq$ & $<$ \\
$C\stackrel{\triangle}{=}\left(>,T\right)$ & $<$ & $\geq$ \\
$C\stackrel{\triangle}{=}\left(\geq,T\right)$ & $\leq$ & $>$ \\
\hline\\
\end{tabular}
\label{tbl:open-pos-nec-rels}
}

@@

Where the constraints $C_1$ and $C_2$ are given in equations \eqref{eq:constraint-c1} and \eqref{eq:constraint-c2}.

\begin{example}
\label{ex:open-time-interval}
Consider an ill-known time interval given by $\left[\FB, Y \right]$. Consider also that in this case, $Y= \left[15/10/2012,3,4 \right] $. Figure \ref{fig:example-open-time-interval} shows the representation for $Y$. The user wants to obtain the possibility and the necessity measures for the $\FB$ part of the interval are:
\end{example}

@@

\vglue13pt
%\begin{table}[htbp]
\tcap{All the possible combination of values for the time interval $\left[X, Y\right]$. The subtypes refer to Table \ref{tbl:time-point-types}.}
\centerline{\small TIME INTERVAL DATA TYPE}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{p{0.15\columnwidth} p{0.15\columnwidth} p{0.5\columnwidth}}\\
\hline
Subtype for $X$  & Subtype for $Y$  & Description\\
\hline
1 or 2 & 1 or 2 &  An ill-known time interval.\\
3 & 3 & An unknown time interval. \\
$\FB$ & 1 or 2 & A left-open time interval.\\
1 or 2 & $\UC$ & A right-open time interval.\\
\hline\\
\end{tabular}
\label{tbl:time-interval-types}
}

% Maybe an example with a database.
% \begin{example}
% 
% \end{example}

\subsection{\label{subsec:fuzzy-allen-relations}Allen's relations}
The Allen's relations can be represented by a Boolean combination of ill-known constraints. The constructs for the constraints are shown in Table \ref{tbl:fuzzy-allen-relations}. The possibility and necessity measures are obtained for each Boolean function as explained in Section \ref{subsec:interval-evaluation-by-ill-known-constraints} in equations \eqref{eq:conjunctive1}-\eqref{eq:negation}.

\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:fuzzy-allen-relations}Constructs of constraints related to their respective Allen relations, as used in the presented work. In this table, the ill-known time interval $J = \left[X, Y\right]$ in a record $r$ has a start point described by possibilistic variable $X$ and an end point described by possibilistic variable $Y$. The crisp time interval in the user's temporal demand is denoted $I$.}
\centerline{\small Relations}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{p{0.2\columnwidth} p{0.8\columnwidth} }\\
\hline
Allen Relation & Construct of constraints\\
\hline
I before J & $C_1 \triangleq \left(<,X\right)$ \\
\hline
I equal J & $\left(C_1 \triangleq \left(\geq,X\right)\right) \wedge \neg \left(C_2 \triangleq \left(\neq,X\right) \right) \wedge \left(C_3 \triangleq \left(\leq,Y\right) \right) \wedge \neg \left(C_4 \triangleq \left(\neq,Y\right)\right)$ \\
\hline
I meets J & $\left(C_1 \triangleq \left(\leq,X\right)\right) \wedge \neg \left(C_2 \triangleq \left(\neq,X\right) \right)$ \\
\hline
I overlaps J & $\left(C_1 \triangleq \left(<,Y\right)\right) \wedge \neg \left(C_2 \triangleq \left(\leq,X\right) \right) \wedge \neg \left(C_3 \triangleq \left(\geq,X\right) \right)$ \\
\hline
I during J & $\left( \left (C_1 \triangleq \left(>,X\right) \right) \wedge \left(C_2 \triangleq \left(\leq,Y\right) \right) \right) \vee \left( \left(C_3 \triangleq \left(\geq,X\right) \right) \wedge  \left(C_4 \triangleq \left(<,Y\right) \right) \right)$ \\
\hline
I starts J & $\left(C_1 \triangleq \left(\geq,X\right) \right) \wedge \neg \left(C_2\triangleq \left(\neq,X\right)\right)$ \\
\hline
I finishes J & $\left(C_1 \triangleq \left(\leq,Y\right) \right) \wedge \neg \left(C_2\triangleq \left(\neq,Y\right)\right)$ \\
\hline\\
\end{tabular}
}
